---
title: "Project DATA 606 Data Proposal"
author: "Guibril Ramde"
date: "2026-04-12"
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data
library(tidyverse)
library(janitor)
library(broom)
library(knitr)
data_url <- "https://raw.githubusercontent.com/japhet125/Project606/refs/heads/main/lifestyle_weight_tracker.csv"
load_data <- read_csv(data_url)

clean_data <- load_data |>
  clean_names()


clean_data2 <- clean_data |>
  filter(
    calories_consumed > 0,
    protein_g > 0,
    carbs_g > 0,
    fat_g > 0
  ) |>
  drop_na()
clean_data2 <- clean_data2 |>
  mutate(
    height_m = height_cm / 100,
    bmi = current_weight_kg / (height_m^2)
  )

clean_data2 |>
  group_by(workout_intensity) |>
  summarise(
    mean_weight_change = mean(weight_change),
    sd_weight_change = sd(weight_change),
    n = n()
  )

glimpse(clean_data2)
dim(clean_data2)
```


### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

Answer: My research question is: Which lifestyle factors (calorie intake, sleep, stress, caffeine, physical activity, and workout intensity) are significantly associated with weight change?


### Cases 

**What are the cases, and how many are there?**
Answer:The cases are the individual observations (rows) in the dataset, where each row represents one tracked lifestyle and weight record. There are initially 55,000 observations but after cleaning and removing NA values we have 52,236 cases in the dataset.



### Data collection 

**Describe the method of data collection.**

Answer: This dataset was obtained from Kaggle as a secondary data source. The file was downloaded as a CSV and then loaded into R using a GitHub raw link. Based on the dataset description, it appears to contain recorded lifestyle and biometric variables related to weight tracking. However, the original data collection and sampling procedures are not fully documented, so the results should be interpreted with caution.

### Type of study 

**What type of study is this (observational/experiment)?**
Answer: This is an observational study because the dataset records lifestyle and health-related variables without assigning treatments or manipulating conditions.


### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

Answer: Lifestyle Weight Tracker Dataset. The link:https://www.kaggle.com/datasets/waddahali/lifestyle-weight-tracker


### Describe your variables?

**Are they quantitative or qualitative**
Answer: the dataset includes both variables quantitative and qualitative:
categorical (qualitative) variables are: 
- gender,
- workout_type, 
- workout_intensity 
numerical (quantitative) varibales are:
- age,
- height_cm,
- initial_weight_kg,
- sleep_hours,
- caffeine_mg,
- calories_consumed,
- protein_g,
- carbs_g,
- fat_g,
- steps,
- temp_c,
- weight_change,
- current_weight_kg

**If you are are running a regression or similar model, which one is your dependent variable?**
Answer: The dependent variable is `weight_change`, which is a numerical (continuous) variable representing the change in weight over time.

### Relevant summary statistics 

**Provide summary statistics for each of the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}
library(gt)
library(kableExtra)
summary_stats <- clean_data2 |>
  summarise(
    n = n(),
    mean_weight_change = mean(weight_change, na.rm = TRUE),
    median_weight_change = median(weight_change, na.rm = TRUE),
    sd_weight_change = sd(weight_change, na.rm = TRUE),
    mean_calories = mean(calories_consumed, na.rm = TRUE),
    mean_sleep = mean(sleep_hours, na.rm = TRUE),
    mean_stress = mean(stress_level, na.rm = TRUE),
    mean_steps = mean(steps, na.rm = TRUE),
    mean_caffeine = mean(caffeine_mg, na.rm = TRUE)
  )

kable(summary_stats, digits = 2, caption = "Summary Statistics")
#kable_styling() |>
  #save_kable("summary_table.png")
summary_stats |>
  gt() |>
 gtsave("summary_table.png")
```
Interpretation: Some variables contain unrealistic negative values and missing values, indicating potential data quality issues which we cleaned in the data preparation before analysis.

##Adding regression model
```{r}
weight_model <- lm(
  weight_change ~ calories_consumed + sleep_hours + stress_level + caffeine_mg + steps +
    workout_intensity,
  data =  clean_data2
)
summary(weight_model)
anova(weight_model)
```

#The overall F-test is statistically significant (p < 0.001), confirming that the model provides a better fit than a model with no predictors.
##There is strong evidence that at least one of the predictor variables is significantly associated with weight change.
##Calories consumed and stress level are the strongest predictors of weight change, as they have very large F-values and extremely small p-values. Caffeine intake and workout intensity are not statistically significant, suggesting they do not meaningfully contribute to predicting weight change in this model.

##Regression output
```{r}
tidy(weight_model) |>
  kable(digits = 4, caption = "Regression Output")

```

##R-squared
```{r}
glance(weight_model) |>
  select(r.squared, adj.r.squared, p.value) |>
  kable(digits = 4, caption = "Model Fit Statistics")


```

##The R-squared value of 0.856 means that approximately 85.6% of the variation in weight change is explained by the lifestyle variables included in the model (calories, sleep, stress, caffeine, steps, and workout intensity. This indicates a strong model fit, suggesting that these variables collectively explain most of the variation in weight change. A high R-squared does not mean all variables are important individually — some predictors (like sleep_hours and caffeine_mg) are not statistically significant based on their p-values.

##The regression model examines whether lifestyle factors such as calorie intake, sleep, stress, caffeine intake, steps, and workout intensity are associated with weight change. The R-squared value shows the proportion of variation in weight change explained by these predictors. A higher R-squared means the model explains more of the differences in weight change, while a low R-squared suggests that other factors not included in the model may also be important.

**Outliers**
```{r}
ggplot(clean_data2, aes(y = calories_consumed)) +
  geom_boxplot() +
  labs(title = "Calories Consumed")
ggsave("Calories_Consumed.png", width = 10, height = 5.6, dpi = 300)
```
```{r}
ggplot(clean_data2, aes(x = weight_change)) +
  geom_histogram(bins = 30, fill = "blue") +
  labs(title = "Distribution of Weight Change")
ggsave("Weight_distribution.png", width = 10, height = 5.6, dpi = 300)
```

Interpretation: The distribution of weight change appears approximately symmetric and roughly bell-shaped, suggesting it may be close to normal.

```{r}
ggplot(clean_data2, aes(x = workout_type, y = weight_change)) +
  geom_boxplot() +
  labs(
    title = "Work Out Type vs Weight Change",
    x = "Workout Type",
    y = "Weight Change"
  )
ggsave("Weight_Change.png", width = 10, height = 5.6, dpi = 300)
```


```{r}
ggplot(clean_data2, aes(x = stress_level, y = weight_change)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Stress Level Vs Weight Change",
    x = "Stress Level",
    y = "Weight Change"
  ) +
  theme_minimal()
ggsave("Stress_Level.png", width = 10, height = 5.6, dpi = 300)
```


```{r}
ggplot(clean_data2, aes(x = sleep_hours, y = weight_change)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Sleep Hours Vs Weight Change",
    x = "Sleep Hours",
    y = "Weight Change"
  ) +
  theme_minimal()
ggsave("Sleep_Hours.png", width = 10, height = 5.6, dpi = 300)
```
```{r}
ggplot(clean_data2, aes(x = sleep_hours, y = weight_change)) +
  geom_bin2d() +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  theme_minimal()
```

```{r}
ggplot(clean_data2, aes(x = calories_consumed, y = weight_change)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Calories Consumed vs Weight Change",
    x = "Calories Consumed",
    y = "Weight Change"
  ) +
  theme_minimal()
ggsave("Weight_Calorie.png", width = 10, height = 5.6, dpi = 300)
```


```{r}
ggplot(clean_data2, aes(x = factor(workout_intensity), y = weight_change)) +
  geom_boxplot() +
  labs(
    title = "Weight Change by Workout Intensity",
    x = "Workout Intensity",
    y = "Weight Change"
  )
ggsave("Workout_Intensity.png", width = 10, height = 5.6, dpi = 300)
```


```{r}
clean_data2 |>
  group_by(workout_intensity) |>
  summarise(
    Q1 = quantile(weight_change, 0.25),
    Median = median(weight_change),
    Q3 = quantile(weight_change, 0.75)
  )
```
##### Hypotheses

Null hypothesis (H0): None of the lifestyle variables are associated with weight change.

Alternative hypothesis (H1): At least one lifestyle variable is associated with weight change.

### Conclusion

This analysis shows that calorie intake and stress level are the strongest predictors of weight change. Steps also have a statistically significant effect, while sleep, caffeine intake, and workout intensity do not show significant effects in the full model.

### Why this matters

Understanding how lifestyle factors influence weight change can help individuals and health professionals make better decisions about diet, stress management, and physical activity.

### Limitations

This is an observational dataset, so causation cannot be established. Additionally, the data collection process is not fully documented, and some variables may be self-reported, which can introduce bias or measurement error.
